# rethinking {#rethinking}

## Resources

- [Statistical Rethinking](https://xcelab.net/rm/statistical-rethinking/) by McElreath

- [rethinking github repo](https://github.com/rmcelreath/rethinking)

## Description

## Environment Setup

```{r setup, results="hide", message=FALSE}
rm(list=ls())

set.seed(123)
options("scipen" = 1, "digits" = 4)

library(tidyverse)
library(datasets)
data(mtcars)

library(rethinking)

```

## Linear Model

### Define Model

```{r mdl1, results="hide"}
# Note the sign change for mu and b, this seems to be a quirk
# of map2stan that it didn't like b ~ dnorm(-0.2, 0.1)
f <- alist(
  mpg ~ dnorm(mu, sigma),
  mu <- a - b * disp,
  a ~ dnorm(25, 10),
  b ~ dnorm(0.2, 0.1),
  sigma ~ dexp(1)
)

# Note the default number of chains = 1, so I'm explicitly setting to 4 here
mdl1 <- map2stan(f,mtcars, chains=4)
```

### Prior Predictive Distribution

```{r mdl1_prior}
# Plot prior predictive distribution
N <- 100

prior_samples <- as.data.frame(extract.prior(mdl1, n=N))

D <- seq(min(mtcars$disp), max(mtcars$disp), length.out = N)

res <- as.data.frame(apply(prior_samples, 1, function(x) x[1] - x[2] * (D-mean(mtcars$disp)))) %>%
  mutate(disp = D) %>%
  pivot_longer(cols=c(-"disp"), names_to="iter") 

res %>%
  ggplot() +
  geom_line(aes(x=disp, y=value, group=iter), alpha=0.2) +
  labs(x="disp", y="prior predictive mpg")
```

### Diagnostics

The _traceplot_ function is the _rethinking_ equivalent to _mcmc_trace_.

```{r mdl1_trace}
traceplot(mdl1@stanfit)
```

The _trankplot_ function is the _rethinking_ equivalent to the _mcmc_rank_overlay_.  

```{r mdl1_trank}
trankplot(mdl1)
```

The _trankplot_ function conveniently also displays the effective sample size (_n_eff_).  But the _precis_ function is another way to get both _n_eff_ and $\hat{R}$.

```{r mdl1_precis}
precis(mdl1, prob=0.95)
```

### Posterior Distribution

The _precis_ function above also displays both the posterior point estimate and credible interval.

### Posterior Predictive Distribution

Finally, I'll check the posterior predictive distribution. The _rethinking_ package includes the _postcheck_ function which displays a plot for posterior predictive checking.

```{r mdl1_ppd, results="hide"}
postcheck(mdl1, window=nrow(mtcars))
```

Alternatively, I can plot the expectation of the posterior predictive distribution (i.e., $\mu$) like I did with _rstanarm_ . The _sim_ function draws samples from the posterior predictive distribution, and the _link_ function returns the linear predictor, possibly transformed by the inverse-link function. In this case, the model is a Gaussian likelihood with an identity link function, so the _sim_ and _link_ functions return identical results.

```{r mdl1_eppd, results="hide"}
newdata <- data.frame(disp=seq(min(mtcars$disp), max(mtcars$disp)))

y_rep <- as.data.frame(t(link(mdl1, data=newdata, n=50))) %>%
  cbind(newdata) %>%
  pivot_longer(cols=starts_with("V"), names_to="grp", values_to="mpg")

y_rep %>%
  ggplot(aes(x=disp, y=mpg)) +
  geom_line(aes(group=grp), alpha=0.2) +
  geom_point(data = mtcars) 
```

## Semi-parametric Model

### Define Model

Setting up the semi-parametric model is a bit more work in the _rethinking_ package.  First, I explicitly create the splines.  The component splines are plotted below.

```{r spline}
library(splines)

num_knots <- 4  # number of interior knots
knot_list <- quantile(mtcars$disp, probs=seq(0,1,length.out = num_knots))
B <- bs(mtcars$disp, knots=knot_list[-c(1,num_knots)], intercept=TRUE)

df1 <- cbind(disp=mtcars$disp, B) %>%
  as.data.frame() %>%
  pivot_longer(-disp, names_to="spline", values_to="val")

# Plot at smaller intervals so curves are smooth
N<- 50
D <- seq(min(mtcars$disp), max(mtcars$disp), length.out = N)
B_plot <- bs(D, 
             knots=knot_list[-c(1,num_knots)], 
             intercept=TRUE)

df2 <- cbind(disp=D, B_plot) %>%
  as.data.frame() %>%
  pivot_longer(-disp, names_to="spline", values_to="val")

ggplot(mapping=aes(x=disp, y=val, color=spline)) +
  geom_point(data=df1) +
  geom_line(data=df2, linetype="dashed")
```

Then I define the model with the splines.  I wasn't able to get this model to work with either the _map2stan_ or _ulam_ functions, so I used _quap_ instead which fits a quadratic approximation.

```{r mdl2, results='hide'}
f <- alist(
  mpg ~ dnorm(mu, sigma),
  mu <- a - B %*% w,
  a ~ dnorm(25, 10),
  w ~ dnorm(0,5),
  sigma ~ dexp(1)
)


mdl2 <- quap(f, data=list(mpg=mtcars$mpg, B=B),
             start=list(w=rep(1, ncol(B)))
            )
```

### Prior Predictive Distribution

### Diagnostics

Since MCMC was not used to fit the model, there are no chain diagnostics to examine. 

### Posterior Distribution

I can still use the _precis_ function to look at the postrior distribution, although there's really no intuitive interpretation for the spline weights.

```{r mdl2_post}
precis(mdl2, depth=2)
```

### Posterior Predictive Distribution

And finally, the posterior predictive distribution:

```{r mdl2_ppd, figures-side, fig.show="hold", out.width="50%", message=FALSE}
mu <- link(mdl2)
mu_mean <- as.data.frame(apply(mu, 2, mean)) %>%
  mutate(disp=mtcars$disp)
colnames(mu_mean) <- c("mpg_ppd", "disp")

mu_PI <- as.data.frame(t(apply(mu,2,PI,0.95))) %>%
  mutate(disp=mtcars$disp)
colnames(mu_PI) <- c("lwr", "upr", "disp")

ggplot() +
  geom_point(data=mtcars, aes(x=disp, y=mpg)) +
  geom_line(data=mu_mean, aes(x=disp, y=mpg_ppd), color="blue") +
  geom_ribbon(data=mu_PI, aes(x=disp, ymin=lwr, ymax=upr), alpha=0.2) +
  labs(title="GAM")

ggplot(mapping=aes(x=disp, y=mpg-mean(mpg)),
              data=mtcars) +
  geom_point()+
  stat_smooth(method="loess",
              level=0.95) +
  labs(title="LOESS")

```

## Extras

```{r}
stancode(mdl1)
```

