# brms {#brms}

## Resources

- [Overview of brms](https://cran.r-project.org/web/packages/brms/vignettes/brms_overview.pdf)

- [Solomon Kurz's translation of _Statistical Rethinking_](https://bookdown.org/content/4857/)

## Description


## Environment Setup

```{r setup, results="hide", message=FALSE}
rm(list=ls())

set.seed(123)
options("scipen" = 1, "digits" = 4)

library(tidyverse)
library(datasets)
data(mtcars)

library(brms)
library(bayesplot)

```

## Linear Model

### Define Model

Like the `rethinking` package, `rstan` doesn't have default priors, so I need to explicitly choose them:

\begin{align*}
  mpg &\sim N(\mu, \sigma^2) \\
  \mu &= a + b*disp \\
  a &\sim N(25,10) \\
  b &\sim N(-0.2, 0.1) \\
  \sigma &\sim Exp(1)
\end{align*}


### Prior Predictive Distribution



### Diagnostics


### Posterior Distribution



### Posterior Predictive Distribution


## Semi-parametric Model

### Define Model

First, I'll define the splines just as I did with the `rethinking` package.  

```{r splines}
library(splines)

num_knots <- 4  # number of interior knots
knot_list <- quantile(mtcars$disp, probs=seq(0,1,length.out = num_knots))
B <- bs(mtcars$disp, knots=knot_list[-c(1,num_knots)], intercept=TRUE)

df1 <- cbind(disp=mtcars$disp, B) %>%
  as.data.frame() %>%
  pivot_longer(-disp, names_to="spline", values_to="val")

# Plot at smaller intervals so curves are smooth
N<- 50
D <- seq(min(mtcars$disp), max(mtcars$disp), length.out = N)
B_plot <- bs(D, 
             knots=knot_list[-c(1,num_knots)], 
             intercept=TRUE)

df2 <- cbind(disp=D, B_plot) %>%
  as.data.frame() %>%
  pivot_longer(-disp, names_to="spline", values_to="val")

ggplot(mapping=aes(x=disp, y=val, color=spline)) +
  geom_point(data=df1) +
  geom_line(data=df2, linetype="dashed")
```

Note: the dashed lines are the splines and the points are the values of the spline at the specific values of `mtcars$disp`; the points are inputs into the `stan` model.



### Prior Predictive Distribution



### Diagnostics



### Posterior Distribution



### Posterior Predictive Distribution



## Semi-parametric Model (Random Walk Prior)

### Define Model

### Prior Predictive Distribution

### Diagnostics

### Posterior Distribution

### Posterior Predictive Distribution


## Session Info

```{r}
sessionInfo()
```

