# rstan {#rstan}

## Resources

- [mc-stan online documentation](https://mc-stan.org/users/interfaces/rstan)

## Description

All of the packages in the previous chapters are running `stan` under the hood. One of the biggest advantages to using `rstan` (or `cmdstanr`) is you get the full power and flexibility of `stan` so you can build models that aren't supported by the other packages. The tradeoff is the syntax is quite different which can be a challenge for those who are only familiar with R.

## Environment Setup

```{r setup, results="hide", message=FALSE}

set.seed(123)
options("scipen" = 1, "digits" = 4)

library(tidyverse)
library(datasets)
data(mtcars)

library(rstan)
library(bayesplot)

# Saves compiled version of model so it only has to be recompiled if the model is changed
rstan_options(auto_write = TRUE)
# Set number of cores
options(mc.cores = parallel::detectCores())
```

## Linear Model

### Define Model

Like the `rethinking` package, `rstan` doesn't have default priors, so I need to explicitly choose them:

\begin{align*}
  mpg &\sim Normal(\mu, \sigma^2) \\
  \mu &= a + b*disp \\
  a &\sim Normal(13.2, 5.3^2) \\
  b &\sim Normal(-0.1, 0.05^2) \\
  \sigma &\sim Exponential(1)
\end{align*}

For a simple linear model there are three sections to the model definition:

1. `data` - This is where the data structures for the known/observed portions of the model (e.g., the number of observations, the number and type of predictors) are defined.

2. `parameters` - This is where the data structures for the parameters to be estimated are defined. For example, the coefficients of the simple linear model belong in this section.

3. `model` - This is where the model (including priors) is defined using the data structures from the previous sections.

```{r mdl1_definition}
# Define model

mdl_code <- '
  data{
    int<lower=1> N;
    vector[N] mpg;
    vector[N] disp;
  }
  parameters{
    real a;
    real b;
    real<lower=0.0> sigma;
  }
  model{
    // Likelihood
    mpg ~ normal(a + b * disp, sigma);
    // Priors
    a ~ normal(13.2, 5.3);
    b ~ normal(-0.1, 0.05);
    sigma ~ exponential(1);
  }
'
```

A couple of comments about the model definition.  

1. For those only familiar with R, it may seem like a lot of extra "stuff" is going on in the `data` and `parameters` sections.  This is because `stan` is written in C++ which is statically typed, unlike R and Python which are dynamically typed.  Essentially what that means is you must define the type of any variable before you use it.

2. The `lower=` (and `upper=` not shown in this example) options define bounds for a variable. The data is checked against the bounds which can detect errors pre-compilation. Generally, bounds are a good idea but aren't required.

Next, populate the data structures from the `data` section and save in a list.

```{r md1_data}
mdl_data <- list(N = nrow(mtcars),
                 mpg = mtcars$mpg,
                 disp = mtcars$disp)
```

And this is the call to fit the model.

```{r mdl1, results='hide'}
# Fit model
mdl1 <- stan(model_code=mdl_code, data=mdl_data, model_name="mdl1")
```

### Prior Predictive Distribution

I could manually construct the prior predictive distribution like I did in \@ref(rethinkingprior).  Instead I'll have `stan` generate the prior predictive distribution which will be useful for more complex models. First, create another model with just the `data` and `generated quantities` section.  The `generated quantities` section mirrors the `model` section except it is now drawing samples from the priors without conditioning on the observed data. Also, in the `stan` call set the sampling algorithm for fixed parameters.

```{r mdl1_prior, results='hide'}
# Plot prior predictive distribution
mdl_prior <- '
  data{
    int<lower=1> N;
    vector[N] disp;
  }
generated quantities{
  real a_sim = normal_rng(13.2, 5.3);
  real b_sim = normal_rng(-0.1, 0.05);
  real sigma_sim = exponential_rng(1);
  real mpg_sim[N] = normal_rng(a_sim + b_sim * disp, sigma_sim);
}
'

N<- 50
D <- seq(min(mtcars$disp), max(mtcars$disp), length.out = N)
mdl_data_prior <- list(N = N, disp=D)

mdl_prior <- stan(model_code=mdl_prior, data=mdl_data_prior, model_name="mdl_prior",
             chains=1, algorithm="Fixed_param")

```

```{r mdl1_prior_plot}
draws <- as.data.frame(mdl_prior) %>%
  head(50)

# Expected value prior predictive distribution
exp_mpg_sim <- apply(draws, 1, function(x) x["a_sim"] + x["b_sim"] * (D-mean(mtcars$disp))) %>%
  as.data.frame() %>%
  mutate(disp = D) %>%
  pivot_longer(-c("disp"), names_to="iter", values_to="mpg") 

ggplot() +
  geom_line(data=exp_mpg_sim, mapping=aes(x=disp, y=mpg, group=iter), alpha=0.2) 
```

### Diagnostics

```{r mdl1_trankplot}
mcmc_rank_overlay(mdl1, pars=c("a", "b", "sigma"))
```

```{r mdl1_summary}
print(mdl1)
```

### Posterior Distribution

The `print` function above displays information about the posterior distributions in addition to `n_eff`.  Alternatively, the `plot` function provides a graphical display of the posterior distributions.

```{r mdl1_post_plot}
plot(mdl1, ci_level=0.89)
```

### Posterior Predictive Distribution

Using the posterior samples, I can plot the expected value of the posterior predictive distribution.

```{r mdl1_ppd}

N<- 50
D <- seq(min(mtcars$disp), max(mtcars$disp), length.out = N)

draws <- as.data.frame(mdl1) %>%
  head(50)

# Expected value posterior predictive distribution
post_pred <- apply(draws, 1, function(x) x["a"] + x["b"]*(D)) %>%
  as.data.frame() %>%
  mutate(disp = D) %>%
  pivot_longer(-c("disp"), names_to="iter", values_to="mpg") 

ggplot() +
  geom_line(data=post_pred, mapping=aes(x=disp, y=mpg, group=iter), alpha=0.2) +
  geom_point(data=mtcars, mapping=aes(x=disp, y=mpg))
```

Note that the expected value of the ppd doesn't include $\sigma$. 

An alternative is to have `stan` automatically generate samples from the posterior predictive distribution by adding a `generated quantities` section to the model (similar to what I did for the prior predictive distribution). 

```{r mdl1_ppd_alternate, results="hide"}
# Define model

mdl_code_ppd <- '
  data{
    int<lower=1> N;
    vector[N] mpg;
    vector[N] disp;
  }
  parameters{
    real a;
    real b;
    real<lower=0.0> sigma;
  }
  transformed parameters{
    // Expected value of posterior predictive
    vector[N] Y_hat = a + b * disp;
  }
  model{
    // Likelihood
    mpg ~ normal(Y_hat, sigma);
    // Priors
    a ~ normal(13.2, 5.3);
    b ~ normal(-0.1, 0.05);
    sigma ~ exponential(1);
  }
  generated quantities{
    // Posterior Predictive
    real mpg_ppd[N] = normal_rng(Y_hat, sigma);
  }
'

# Fit model
mdl1_ppd <- stan(model_code=mdl_code_ppd, data=mdl_data)

draws <- as.data.frame(mdl1_ppd)

# 95% credible interval for expected value of ppd
Eppd <- draws %>% select(starts_with("Y_hat")) %>%
  apply(2, function(x) quantile(x, probs=c(0.025, 0.5, 0.975))) %>%
  t() %>%
  as.data.frame() %>%
  mutate(disp = mtcars$disp)

# 95 credible interval for ppd
ppd <- draws %>% select(starts_with("mpg_ppd")) %>%
  apply(2, function(x) quantile(x, probs=c(0.025, 0.5, 0.975))) %>%
  t() %>%
  as.data.frame() %>%
  mutate(disp=mdl_data$disp)
  
ggplot() +
  geom_line(data=Eppd, mapping=aes(x=disp, y=`50%`)) +
  geom_ribbon(data=ppd, mapping=aes(x=disp, ymin=`2.5%`, ymax=`97.5%`), 
              alpha=0.5, fill="lightblue") +
  geom_ribbon(data=Eppd, mapping=aes(x=disp, ymin=`2.5%`, ymax=`97.5%`), 
              alpha=0.5, fill="dodgerblue") +
  geom_point(data=mtcars, mapping=aes(x=disp, y=mpg)) +
  labs(x="disp", y="mpg")
```

The darker blue area is a 95% credible interval for the expected value of the posterior predictive distribution and the lighter blue area is the 95% credible interval for the posterior predictive distribution.

And we can also plot the density overlay using the posterior predictive draws.

```{r}
yrep <- draws %>%
  head(50) %>%
  select(starts_with("mpg_ppd")) %>%
  as.matrix()

ppc_dens_overlay(mtcars$mpg, yrep)
```

## Semi-parametric Model

### Define Model

First, I'll define the splines just as I did with the `rethinking` package.  

```{r splines}
library(splines)

num_knots <- 4  # number of interior knots
knot_list <- quantile(mtcars$disp, probs=seq(0,1,length.out = num_knots))
B <- bs(mtcars$disp, knots=knot_list[-c(1,num_knots)], intercept=TRUE)

df1 <- cbind(disp=mtcars$disp, B) %>%
  as.data.frame() %>%
  pivot_longer(-disp, names_to="spline", values_to="val")

# Plot at smaller intervals so curves are smooth
N<- 50
D <- seq(min(mtcars$disp), max(mtcars$disp), length.out = N)
B_plot <- bs(D, 
             knots=knot_list[-c(1,num_knots)], 
             intercept=TRUE)

df2 <- cbind(disp=D, B_plot) %>%
  as.data.frame() %>%
  pivot_longer(-disp, names_to="spline", values_to="val")

ggplot(mapping=aes(x=disp, y=val, color=spline)) +
  geom_point(data=df1) +
  geom_line(data=df2, linetype="dashed")
```

Note: the dashed lines are the splines and the points are the values of the spline at the specific values of `mtcars$disp`; the points are inputs into the `stan` model.

```{r mdl1_gam, results="hide"}
# Define model

mdl_code <- '
  data{
    int<lower=1> N;
    int<lower=1> num_basis;
    vector[N] mpg;
    vector[N] disp;
    matrix[N, num_basis] B;
  }
  parameters{
    real a;
    real<lower=0.0> sigma;
    vector[num_basis] w;
  }
  transformed parameters{
    vector[N] Y_hat = a + B*w;
  }
  model{
    // Likelihood
    mpg ~ normal(Y_hat, sigma);
    // Priors
    a ~ normal(25, 10);
    sigma ~ exponential(1);
    w ~ normal(0, 5);
  }
  generated quantities{
    // Posterior Predictive
    real mpg_ppd[N] = normal_rng(Y_hat, sigma);
  }
'

mdl_data <- list(N=nrow(mtcars),
                 num_basis=ncol(B),
                 B=B,
                 mpg = mtcars$mpg,
                 disp = mtcars$disp)

# Fit model
mdl1_gam <- stan(model_code=mdl_code, data=mdl_data)
```

### Prior Predictive Distribution

```{r mdl1_gam_prior, results="hide"}
# Define model

mdl2 <- '
  data{
    int<lower=1> N;
    int<lower=1> num_basis;
    //vector[N] mpg;
    //vector[N] disp;
    matrix[N, num_basis] B;
  }
  generated quantities{
    real a_sim = normal_rng(25, 10);
    real sigma_sim = exponential_rng(1);
    real mpg_sim[N];
    vector[N] Y_hat;
    vector[num_basis] w_sim;
    for (i in 1:num_basis)
      w_sim[i] = normal_rng(0,5);
    Y_hat = a_sim + B * w_sim;
    mpg_sim = normal_rng(Y_hat, sigma_sim);
}
'

mdl_gam_prior <- stan(model_code=mdl2, 
                      data=list(N=N, 
                                num_basis=ncol(B_plot),
                                B=B_plot),
                      chains=1, algorithm="Fixed_param")

draws <- as.data.frame(mdl_gam_prior) %>%
  head(50)

# Expected value prior predictive distribution
exp_mpg_sim <- apply(draws, 1, function(x) {
  x["a_sim"] + B_plot %*% x[grepl("w", names(x))]
}) %>%
  as.data.frame() %>%
  mutate(disp = D) %>%
  pivot_longer(-c("disp"), names_to="iter", values_to="mpg") 

# 95% interval prior predictive distribution
mpg_sim <- as.data.frame(mdl_gam_prior) %>% select(starts_with("mpg")) %>%
  apply(2, function(x) quantile(x, probs=c(0.025, 0.5, 0.975))) %>%
  t() %>%
  as.data.frame() %>%
  mutate(disp = D)

ggplot() +
  geom_line(data=exp_mpg_sim, mapping=aes(x=disp, y=mpg, group=iter), alpha=0.2) +
  geom_ribbon(data=mpg_sim, mapping=aes(x=disp, ymin=`2.5%`, ymax=`97.5%`), 
              alpha=0.5, fill="lightblue")
```

### Diagnostics

```{r mdl1_gam_trankplot}
# Note that bayesplot methods support tidy selection of parameters
mcmc_rank_overlay(mdl1_gam, pars=vars(a, sigma, starts_with("w[")))
```

```{r mdl1_gam_summary}
# This is the print.stanfit method and pars must be a character vector
print(mdl1_gam, pars=c("a", "sigma", "w"))
```

### Posterior Distribution

```{r mdl1_gam_post}
plot(mdl1_gam, pars=c("a", "sigma", "w"))
```

### Posterior Predictive Distribution

```{r mdl1_gam_ppd}

# 95% credible interval expected value of posterior predictive
Eppd <- as.data.frame(mdl1_gam) %>% select(starts_with("Y_hat")) %>%
  apply(2, function(x) quantile(x, probs=c(0.025, 0.5, 0.975))) %>%
  t() %>%
  as.data.frame() %>%
  mutate(disp = mtcars$disp)

# 95% credible interval posterior predictive
ppd <- as.data.frame(mdl1_gam) %>% select(starts_with("mpg")) %>%
  apply(2, function(x) quantile(x, probs=c(0.025, 0.5, 0.975))) %>%
  t() %>%
  as.data.frame() %>%
  mutate(disp = mtcars$disp)

ggplot() +
  geom_line(data=Eppd, mapping=aes(x=disp, y=`50%`)) +
  geom_ribbon(data=ppd,
              mapping=aes(x=disp, ymin=`2.5%`, ymax=`97.5%`), 
              alpha=0.5, fill="lightblue") +
    geom_ribbon(data=Eppd, mapping=aes(x=disp, ymin=`2.5%`, ymax=`97.5%`), 
              alpha=0.5, fill="dodgerblue") +
  geom_point(data=mtcars, mapping=aes(x=disp, y=mpg)) +
  labs(x="disp", y="mpg")
```

## Semi-parametric Model (Random Walk Prior)

### Define Model

One challenge with splines is choosing the number of knots.  For the previous model, I tried several values for `num_knots` until settling on 4.  However, there is a `stan` case study for splines that uses a novel prior which addresses this issue.  The details are [here](https://mc-stan.org/users/documentation/case-studies/splines_in_stan.html).

For this example, I will set `num_knots=20` and then fit models with and without the random walk prior. This is an example where the `rstan`'s flexibility is an advantage because it would be difficult or impossible to specify the random walk prior in the other packages.

```{r results="hide"}
library(splines)

num_knots <- 20  # number of interior knots
knot_list <- quantile(mtcars$disp, probs=seq(0,1,length.out = num_knots))
B <- bs(mtcars$disp, knots=knot_list[-c(1,num_knots)], intercept=TRUE)

# Define model with smoothing prior
mdl_smooth_code <- '
  data{
    int<lower=1> N;
    int<lower=1> num_basis;
    vector[N] mpg;
    vector[N] disp;
    matrix[N, num_basis] B;
  }
  parameters{
    real a;
    real<lower=0.0> sigma;
    vector[num_basis] w_raw;
    real<lower=0.0> tau;
  }
  transformed parameters{
    vector[num_basis] w;
    vector[N] Y_hat;
    w[1] = w_raw[1];
    for (i in 2:num_basis)
      w[i] = w[i-1] + w_raw[i]*tau;
    Y_hat = a + B*w;
  }
  model{
    // Likelihood
    mpg ~ normal(Y_hat, sigma);
    // Priors
    a ~ normal(25, 10);
    sigma ~ exponential(1);
    w_raw ~ normal(0, 1);
    tau ~ normal(0,1);
  }
  generated quantities{
    real mpg_ppd[N] = normal_rng(a + B*w, sigma);
  }
'

mdl_data <- list(N=nrow(mtcars), 
                 num_basis=ncol(B),
                 B=B,
                 mpg = mtcars$mpg,
                 disp = mtcars$disp)

# Fit model with smoothing prior
mdl2_gam_smooth <- stan(model_code=mdl_smooth_code, data=mdl_data,
                 control=list(adapt_delta=0.99))

# Fit model without smoothing prior
mdl2_gam <- stan(model_code = mdl_code, data=mdl_data,
                 control=list(adapt_delta=0.99))
```

### Diagnostics

```{r fig.height=8}
mcmc_rank_overlay(mdl2_gam_smooth, pars=vars("a", "sigma", starts_with("w")))
```


```{r}
print(mdl2_gam_smooth, pars=c("a", "sigma", "w"))
```

### Posterior Distribution

```{r figures-side, fig.show="hold", out.width="50%", message=FALSE}
plot(mdl2_gam_smooth, pars=c("a", "sigma", "w"))
plot(mdl2_gam, pars=c("a", "sigma", "w"))
```

### Posterior Predictive Distribution

```{r}
draws <- as.data.frame(mdl2_gam) %>% head(100)

Epost_pred <- draws %>% select(starts_with("Y_hat")) %>%
  apply(2, function(x) quantile(x, probs=c(0.055, 0.5, 0.945))) %>%
  t() %>%
  as.data.frame() %>%
  mutate(disp = mtcars$disp)

draws_smooth <- as.data.frame(mdl2_gam_smooth) %>% head(100)

Epost_pred_smooth <- draws_smooth %>% select(starts_with("Y_hat")) %>%
  apply(2, function(x) quantile(x, probs=c(0.055, 0.5, 0.945))) %>%
  t() %>%
  as.data.frame() %>%
  mutate(disp = mtcars$disp)

rbind(Epost_pred %>% select(c("disp", `50%`)) %>% mutate(type="without smoothing prior"),
      Epost_pred_smooth %>% select(c("disp", `50%`)) %>% mutate(type="with smoothing prior")) %>%
  ggplot() +
  geom_line( mapping=aes(x=disp, y=`50%`, linetype=type), color="blue" ) +
  geom_point(data=mtcars, mapping=aes(x=disp, y=mpg)) +
  labs(y="mpg")

```

The plot above shows that even with a large number of knots (in this case 20), the model with the smoothing prior significantly reduces over-fitting when compared to the model without the smoothing prior.

## Session Info

```{r}
sessionInfo()
```

